Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/abstract.tex
done with abstract

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we first symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  %
  Second, we prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} input types, there exist
  inhabitants that can make the function go wrong.
  %
  Third, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Fourth, we evaluate our approach on two data sets
  comprising more than 5000 ill-typed student programs.
  Our technique is able to generate witnesses for
  nearly 90\% of the programs, and our reduction graph
  is yields small counterexamples for 90\% of the witnesses.
\end{abstract}

>>>>

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we first symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} input types, there exist
  inhabitants that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate our approach on two data sets
  comprising more than 4500 ill-typed student programs.
  Our technique is able to generate witnesses for
  nearly 90\% of the programs, and our reduction graph
  is yields small counterexamples for 90\% of the witnesses.
\end{abstract}

### 2015-11-20 paper/pldi16-submission/evaluation.tex
done with abstract

We answer these questions on two sets of known-bad programs, \ie
programs that were rejected by the \ocaml compiler because of a type
error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at UC San Diego. 
%
We recorded each interaction with the \ocaml top-level system over the
course of the first three assignments (IRB \#140608), from which we
extracted 4,407 distinct, ill-typed \ocaml programs.
%
The second dataset -- widely used in the literature -- comes from a
similar course at the University of
Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
ill-typed programs.

>>>>

We answer these questions on two sets of known-bad programs, \ie
programs that were rejected by the \ocaml compiler because of a type
error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at UC San Diego.
%
We recorded each interaction with the \ocaml top-level system over the
course of the first three assignments (IRB \#140608), from which we
extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
%
The second dataset -- widely used in the literature -- comes from a
similar course at the University of
Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
ill-typed programs.

### 2015-11-20 paper/pldi16-submission/introduction.tex
done with abstract

% \paragraph{Problem}
Type errors are a common stumbling block for students trying to learn
typed functional languages like \ocaml and \haskell.
%
The compiler imposes a strict discipline that prevents novices from
running their programs until the type-checker has been appeased.
%
In effect this prioritizes the \emph{static} semantics of the language
over the \emph{dynamic} semantics.
%
This is fine for expert users who are already familiar with the language
and type system, but seems backwards for novices, who are typically
introduced to the dynamic semantics first.

Consider the following ill-typed @fac@ function, where we return
@true@ in the base case instead of @1@.
%
\begin{figure}[h!]
\begin{code}
  let rec fac n =
    if n <= 0 then
      true
    else
      n * fac (n-1)
\end{code}
\caption{Ill-typed \texttt{fac} function that mistakenly
  returns \texttt{true} in the base case instead of \texttt{1}.}
\label{fig:factorial}
\end{figure}

\ocaml's response is to tell us that on line 5, the recursive call to
@fac@ returns a @bool@, but we needed an @int@ for the
multiplication.
%
\begin{verbatim}
File "fac.ml", line 5, characters 8-23:
Error: This expression has type bool
       but an expression was expected
       of type int
\end{verbatim}
%
\ocaml is correct, but we may ask \emph{why} @fac@ returns a
@bool@ instead of an @int@.

Contrast this \emph{static} type error to the \emph{dynamic} type error
that arises in the equivalent \ruby function, when we call
@fac(1)@.
%
\begin{verbatim}
fac.rb:5:in `*': true can't be coerced
                 into Fixnum (TypeError)
  from fac.rb:5:in `fac'
  from fac.rb:10:in `<main>'
\end{verbatim}
%
The same basic information is there, @*@ was given a @bool@ (@true@)
when it expected an @int@ (@FixNum@), but since the error occurred at
runtime, \ruby is able to provide a stack trace that provides important
contextual information about the error.
%
More importantly, being able to \emph{run} the bad program means we can
also test it on other inputs, \eg @0@, or we can load it into a debugger
and step through the execution to see concretely how the error arises.

To make matters worse, it is well known that type inference algorithms
based on unification have a penchant for \emph{reporting} errors far
from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much work has been done on pinpointing the \emph{source} of a type
error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

% \paragraph{Contributions}
In this paper we propose exposing \emph{static} type errors for the
\emph{runtime} crashes that they are.
%
Our first contribution is an algorithm for searching for
\emph{witnesses} to type errors, \ie input vectors that cause a program
to crash with a type error.
%
This is not as trivial as it sounds.
%
While one can indeed execute an \ocaml program without type-checking it
-- the dynamic semantics do not depend on the static semantics -- what
types of values should one pass to the program?
%
We do not want to generate \emph{spurious} witnesses, inputs that the
type-checker would have excluded, so we introduce a semantics for
executing \ocaml programs with ``holes'', values whose type we do not
yet know.
%
% A hole can be passed around through the program until it is demanded by
% a primitive operation, \eg addition, function application, branching,
% etc., at which point we know from the context what type it must have,
% and can instantiate it with a concrete value.
%
Our semantics \emph{conservatively} instantiates holes with concrete
values; if we can find a witness to a type error then there is no valid
typing for the program.

Given a witness to a type error, the prospective \ocaml student may
still be at a loss.
%
The standard \ocaml interpreter and debugging infrastructure expect
well-typed programs, so they cannot be used to investigate \emph{how}
the witness causes the program to crash.
%
Thus, our second contribution is an interactive visualization of the
execution of \ocaml programs (well-typed or not).
%
We present the execution as a \emph{reduction graph}, which records
all of the small-step reductions in the context in which they occur.
%
%  in which the nodes
% are expressions and the edges indicate either a single-step reduction or
% a sub-term relation.
%
This graph is easy to generate during the execution and admits many of
the standard debugging steps, \eg ``step forward'', ``step into'', and
``step over'', as well as backwards motion and jumping to the next (or
previous) function call.
%
The visualization initially shows the source term (\ie the witness) and
the stuck term (or the final value), and allows the user to iteratively
expand the computation to expose intermediate steps, by selecting an
expression and choosing a traversal strategy (\eg ``step forward'').
%
This process of iterative expansion of the trace ensures that the
context of intermediate steps is not lost; whereas traditional debuggers
force one to move between statements, in our visualization you never
lose sight of where you came from.

\paragraph{Outline}
We begin the rest of the paper with an overview of our search algorithm
and trace visualization (\S~\ref{sec:overview}).
%
% Then, we formalize our algorithm and prove it sound
% (\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
% student programs (\S~\ref{sec:eval}).
% %
% Next, we extend the search algorithm to collect the reduction graph and
% formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
% complexity of these traces on our benchmarks
% (\S~\ref{sec:evaluation-nanomaly}).
Then, we formalize our algorithm and prove it sound
(\S~\ref{sec:searching-witness}).
%
Next, we extend the algorithm to collect a reduction graph and formalize
a set of traversals over the graph (\S~\ref{sec:explaining}).
%
Finally, we evaluate our search algorithm and trace generation on a
large set of real student programs (\S~\ref{sec:evaluation}).
%
% Finally, we present the results of a preliminary user study on the full
% system (\S~\ref{sec:user-study}).

>>>>

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@).
and so \ocaml's responds with the error message:
%
\begin{verbatim}
  File "fac.ml", line 5, characters 8-23:
  Error: This expression has type bool
         but an expression was expected
         of type int
\end{verbatim}

This message is makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who is still to develop such a
mental model.
%
To make matters worse, unification based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\begin{code}
let rec fac n =
  if n <= 0 then
    true
  else
    n * fac (n-1)
\end{code}
\end{minipage}
\hspace{0.3in}
\begin{minipage}{.3\linewidth}
  \includegraphics[height=2.5in]{fac-overview.png}
\end{minipage}
\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}}
\label{fig:factorial}
\end{figure}

In this paper we propose a new approach to explaining type
\emph{static} type errors: by \emph{dynamically} witnessing
how exactly an ill-typed program goes wrong.
%
We have developed \toolname\ an interactive tool which uses
the source of the ill-typed function to automatically synthesize
the result on the right in Figure~\ref{fig:factorial}, which
shows how the recursive calls reduce to a configuration where
the program ``goes wrong'' \ie\ the @int@ value @1@ is to be
multiplied with the @bool@ value @true@.
We achieve this via three concrete contributions.

\paragraph{1. Finding Witnesses}
Our first contribution is an algorithm for searching for
\emph{witnesses} to type errors, \ie\ inputs that cause a
program to go wrong~(\S~\ref{sec:semantics}).
%
This problem is surprisingly tricky when we cannot rely on
static type information.
%
In particular, we must avoid the trap of \emph{spurious} inputs
that cause irrelevant problems that would be avoided by picking
values of a different, relevant type.
%
We solve this problem by developing a novel operational semantics
that corresponds to executing the program with \emph{holes}
values whose type is unknown.
%
Our semantics \emph{conservatively} instantiates holes with concrete
values; thereby dynamically inferring the type of the input type
until the point the program goes wrong.
%
We prove that our procedure synthesizes \emph{general witnesses},
which means, intuitively, that if a witness is found for a given
ill-typed function, then \emph{for all} input types, there exist
inhabitants that can make the function go wrong.

Given a witness to a type error, the novice may still be at a loss.
%
The standard \ocaml\ interpreter and debugging infrastructure expect
well-typed programs, so they cannot be used to investigate \emph{how}
the witness causes the program to crash.
%
More importantly, the execution itself may be quite long and may contain
details not relevant to the actual error.

\paragraph{2. Visualizing Witnesses}
Our second contribution is a novel interactive visualization of the
execution of \ocaml\ programs well-typed or not~(\S~\ref{sec:interactive}).
%
We extend the semantics to also build a \emph{reduction graph}
which records all of the small-step reductions and the context
in which they occur.
%
The graph lets us build a visualization that shows a sequence of
steps from the source witness to the stuck term and allows the user
to interactively expand the  computation to expose intermediate steps,
by selecting an expression and choosing a traversal strategy.
%
The strategies include many of the standard debugging moves, \eg\
stepping \emph{forward} or \emph{into} or \emph{over} calls as well
stepping or jumping \emph{backward} to understand how a particular
value was created, while preserving a context of the intermediate
steps that allow the user to keep track of a term's provenance.

% This process of iterative expansion of the trace ensures that
% the context of intermediate steps is not lost; whereas traditional
% debuggers force one to move between statements, in our visualization
% you never lose sight of where you came from.

Obviously, the problem of finding witnesses is undecidable.
In fact, due to the necessarily conservative nature of static
typing, there may not even exist. Thus, our heuristic approach
is only useful if it can find \emph{compact} witnesses for
\emph{real-world} programs.

\paragraph{3. Evaluating Witnesses}
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:eval}).
%
The first is the \uwbench\ data set~\cite{lerner_searching_2007}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for nearly 90\% of the programs, in under a second the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with fewer than 5 steps for 60\%
of the programs, and under 10 steps for 90\% of the programs.

In short, our results show that at least for beginners, ill-typed
programs \emph{do} go wrong, in the vast majority of cases, which,
in turn, opens the door to a completely different, dynamic way
to explain, understand, and appreciate the benefits of static typing.

%% Then, we formalize our algorithm and prove it sound
%% (\S~\ref{sec:searching-witness}).
%% %
%% Next, we extend the algorithm to collect a reduction graph and formalize
%% a set of traversals over the graph (\S~\ref{sec:explaining}).
%% %
%% Finally, we evaluate our search algorithm and trace generation on a
%% large set of real student programs (\S~\ref{sec:evaluation}).
%%
%% \paragraph{Outline}
%% We begin the rest of the paper with an overview of our search algorithm
%% and trace visualization (\S~\ref{sec:overview}).

% Then, we formalize our algorithm and prove it sound
% (\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
% student programs (\S~\ref{sec:eval}).
% %
% Next, we extend the search algorithm to collect the reduction graph and
% formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
% complexity of these traces on our benchmarks
% (\S~\ref{sec:evaluation-nanomaly}).
%
% Finally, we present the results of a preliminary user study on the full
% system (\S~\ref{sec:user-study}).
